---
title: "Neurospora Broad Community Annotation Project: citations.json"
output: html_document
---

Parse the Broad Community Annotation Project dumps to tab files for easy loading into FungiDB

All gene IDs correspond to version 7 

### Load dataset into R
#### Set working directory
```{r, eval=FALSE}
setwd("~/Working/fungidb/fungidb_current/Project_Neurospora_Broad/")
```

#### Download file citations.json
Source: Data dumps from Broad (201402) archived by Jason Stajich 
```{r, eval=FALSE}
# Create data folder
if (!file.exists("data")) {
  dir.create("data")
}
### Download files
fileUrl <- "https://github.com/fungidb/neurospora/blob/master/Community_annotations/curator_raw/Broad_2014_02/citations.json.gz"
download.file(fileUrl, destfile = "./data/citations.json.gz", method = "curl" )
```

#### Parse _citations.json_ to valid json format
This json file (citations.json) does not validate (multiple errors). It needs fixing before loading into R. Validated using [JSON validator](http://jsonformatter.org/) 

Use the script below to:

* remove 'data =' in the first line and the ',' in the next-to-last line 
* change all the single quotes (') in the key names for double quotes (")
* Replace unquoted None (: None) by double quoted None (: "None")
* Remove the bell character present in one of the lines.

Script to parse the json file to a valid json format:
```{r, eval=FALSE}
arg1 <- "-a ./data/citations.json"
arg2 <- "-o citations"
cmd <- paste("perl", "./scripts/parse_broad_dumps_to_validated_json.pl", arg1, arg2)
system(cmd) 
```

#### Read json file:
```{r, eval=FALSE}
# Read json data
library(jsonlite)
library(tidyr)
library(dplyr)
citations <- fromJSON("./data/citations_parsed.json")

# Unnest the citation column. 
# The citation column has nested information. Entries contain a list with multiple dataframes as elements
citations.wide <- unnest(citations)
# Change column names
names(citations.wide) <- c("srcgrp","geneid","evidencecode","pubmedid",
                                     "publication","termid","termname","termnamespace"
                                     ,"termontology","toolevidence")
```

### Types of annotations included in _citations.json_
Filtering by "termontology", there are 5 types of annotations:

* NC Pathway
* NC Reaction
* Functional Catalogue
* Gene Ontology
* None (Publications)

```{r, eval = FALSE}
unique(citations.wide$termontology)
```

### 1. NC Pathway and NC Reaction Annotations 
We only need the EC numbers for loading into FungiDB. The pathways and reactions will be assigned in FungiDB according to EC number:FungiCyc mappings. 

Filter NC Pathway and NC Reaction annotations to tab files for reference. Notes:
* There are reactions with no EC number (transporters).
* There is a reference number in the publication field, it is another geneid from Broad. Confirmed that they correspond to version 7 IDs.

```{r, eval=FALSE}
# Filter NC Pathway annotations and write to file
citations.pathway <- citations.wide[citations.wide$termontology == "NC Pathway", 
                                    c("geneid","termid","termname") ]
names(citations.pathway) <- c("geneid","pathwayid","pathwayname")
write.table(citations.pathway, file = "./tidy/citations_nc_pathways.tsv", quote = FALSE, sep = "\t", row.names = FALSE)
# Filter NC Reaction annotations and write to file
citations.reaction <- citations.wide[citations.wide$termontology == "NC Reaction", 
                                     c("geneid","termid","termname") ]
names(citations.reaction) <- c("geneid","reactionname","ecnumber")
citations.reaction$ecnumber <- gsub("# ", ":", citations.reaction$ecnumber, fixed = TRUE)
write.table(citations.reaction, file = "./tidy/citations_nc_reactions.tsv", quote = FALSE, 
            sep = "\t", row.names = FALSE)
```

Filter EC numbers for loading, one row per geneid:ec_number pair

```{r, eval = FALSE}
withecnumber <- which(grepl("^EC:", citations.reaction$ecnumber, perl = TRUE))
citations.ecnumbers <- citations.reaction[withecnumber, c("geneid","ecnumber")] 
# Remove duplicates (2150/3638)
citations.ecnumbers <- unique(citations.ecnumbers)
write.table(citations.ecnumbers, file = "./tidy/citations_ec_numbers.tsv", quote = FALSE, 
            sep = "\t", row.names = FALSE)
```

### 2. Functional Catalogue Annotations
Annotations based on the Functional Catalogue (FunCat). No evidences provided, are they manual or automatic annotations? => Seem automatic from MIPS

They could be loaded as annotator notes or mapped to GO terms and loaded (there is a mapping file available at GOC).

FunCat annotations from NC12 are in schema 2.1. The mapping file funcat2go uses schema 2.0. GO mappings have to be updated to schema 2.1 before parsing the GO terms.

Parse FunCat annotations to tab file
```{r, eval = FALSE}
# Filter FunCat annotations (12040)
citations.funcat <- citations.wide[citations.wide$termontology == "Functional Catalogue", c("geneid","termid","termname")]
names(citations.funcat) <- c("geneid", "funcatid","funcattermname")
## Remove genes with no assigned category (11710)[330]
citations.funcat <- citations.funcat[citations.funcat$funcattermname != "CLASSIFICATION NOT YET CLEAR-CUT",]
## Remove duplicated rows (5818)
citations.funcat <- unique(citations.funcat)
# Add publication reference and evidence code (automatic)
citations.funcat$pubmedid <- rep("PMID:15486203", length(citations.funcat$geneid))
citations.funcat$evidencecode <- rep("IEA", length(citations.funcat$geneid))
## Order by geneid and write to file
citations.funcat <- citations.funcat[order(citations.funcat$geneid),]
write.table(citations.funcat, "./tidy/citations_funcat.tsv", quote = FALSE, sep = "\t", row.names = FALSE)
```

Map FunCat annotations to GO terms
```{r, eval = FALSE} 
## Download FunCat to GO terms mappings
fileUrl <- "http://geneontology.org/external2go/mips2go"
download.file(fileUrl, destfile = "./data/mips2go_mappings.txt", method = "curl")

## Load funcat2go mappings, some terms do not have mapping to GO terms, add "NA" (fill and na.strings) (1433)
funcat2go <- read.table("./data/mips2go_mappings.txt", header = FALSE, sep = ";",
                        skip = 10, quote = "", colClasses = c("character", "character"), 
                        fill = TRUE, col.names = c("funcatid","gotermid"), na.strings=c("","NA"))
## Remove extra information from the funcat id term column, the termid is enough for mapping
funcat2go$funcatid <- gsub("MIPS_funcat:(.+?)\\s.+$","\\1",funcat2go$funcatid, perl = TRUE)

## Parse GO mappings to schema 2.1
### Load schema 2.0 to 2.1 mappings (1501)
fileUrl2 <- "ftp://ftpmips.gsf.de/fungi/1_FunCat_scheme/MappingFunCat2.0_2.1.txt"
download.file(fileUrl2, "./data/MappingFunCat2.0_2.1.txt", method = "curl")
funcat20to21 <- read.table("./data/MappingFunCat2.0_2.1.txt", skip = 3, header = FALSE, 
                             sep = ">", fill = TRUE, na.strings=c("","NA"), 
                             col.names = c("termid1", "termid2"), 
                             colClasses = c("character","character"))
### Remove extra information from the first column (keys), the termid is enough for mapping
funcat20to21$termid1 <- gsub("^(.+?)\\s.+$","\\1", funcat20to21$termid1, perl = TRUE)

### Fill second column for terms with no changes in id
    # Example with changes "01.02.01.0" > "01.02.02.01
    # Example with no changes: "01.01.03.01"

for (i in 1:length(funcat20to21$termid1)){
    if( is.na(funcat20to21$termid2[i])){
        funcat20to21$termid2[[i]] <- funcat20to21$termid1[[i]]
    }
}

write.table(funcat20to21, "./tidy/funcat2.0_to_funcat2.1_mappings.tsv", quote = FALSE, sep = "\t", row.names = FALSE)

### Transform to data.tables for easy mapping
funcat20to21.table <- data.table(funcat20to21) # mappings 2.0 to 2.1
funcat2go.table <- data.table(funcat2go) # mappings FunCat 2.0 to GO terms
setkey(funcat2go.table, funcatid)
setkey(funcat20to21.table, termid1)

### Update GO term mappings to FunCat 2.1 (1433)
funcat2go.table[funcat20to21.table, funcatid := i.termid2]

### Remove NAs and duplicates (for FunCat terms with multiple GO values only the first 
    # GO term will be stored, usually the most general (!00 check))
funcat2go.table <- na.omit(funcat2go.table) # (966)
funcat2go.table <- unique(funcat2go.table, by = "funcatid") # (811)
write.table(funcat2go.table, "./tidy/funcat_to_go_mappings_schema2.1.tsv", quote = FALSE, sep = "\t", row.names = FALSE)

## Map GO terms to NC FunCat 2.1
funcat.table <- data.table(citations.funcat[,c("geneid","funcatid")])

setkey(funcat.table, funcatid)
funcat.table[funcat2go.table, gotermid := i.gotermid]
sum(is.na(funcat.table$gotermid)) # => 861 NAs of 5818 entries 

## Write file with FunCat annotations and GO term mappings, one row per pair gene:term
citations.funcat.go <- as.data.frame(na.omit(funcat.table)) # (4957)
citations.funcat.go <- citations.funcat.go[order(citations.funcat.go$geneid), ]
citations.funcat.go$pubmedid <- rep("PMID:15486203", length(citations.funcat.go$geneid))
write.table(citations.funcat.go, "./tidy/citations_funcat_go_mappings.tsv", quote = FALSE, sep = "\t", row.names = FALSE)

## Write file with FunCat annotations with no GO mapping
citations.funcat.nogo <- as.data.frame(funcat.table[is.na(funcat.table$gotermid), ]) # (861)
citations.funcat.nogo <- citations.funcat.nogo[order(citations.funcat.nogo$geneid),]
citations.funcat.nogo$pubmedid <- rep("PMID:15486203", length(citations.funcat.nogo$geneid))
write.table(citations.funcat.nogo, "./tidy/citations_funcat_nogo_mappings.tsv", quote = FALSE, sep = "\t", row.names = FALSE)

# Write to GAF format. Using IEA and FunCat publication as reference
citations.funcat.go <- read.table("./tidy/citations_funcat_go_mappings.tsv", sep = "\t", 
                                  header = TRUE)
funcat.gaf <- data.frame(db = rep("EuPathDB", length(citations.funcat.go$geneid)), 
                          dbobjectid = citations.funcat.go$geneid, # geneid
                          dbobjectsymbol = rep(NA, length(citations.funcat.go$geneid)), # gene name
                          qualifier = rep(NA, length(citations.funcat.go$geneid)),
                          goid = citations.funcat.go$gotermid,
                          dbreference = citations.funcat.go$pubmedid,
                          evidencecode = rep("IEA", length(citations.funcat.go$geneid)),
                          withfrom = rep(NA, length(citations.funcat.go$geneid)), # none
                          aspect = citations.funcat.go$termnamespace,
                          dbobjectname = rep(NA, length(citations.funcat.go$geneid)), # product
                          dbobjectsynonym = rep(NA, length(citations.funcat.go$geneid)), # name aliases
                          dbobjecttype = rep("gene", length(citations.funcat.go$geneid)),
                          taxon = rep("taxon:367110", length(citations.funcat.go$geneid)),
                          date = rep("20140201", length(citations.funcat.go$geneid)),
                          assignedby = rep("MIPS_funcat", length(citations.funcat.go$geneid)),
                          annotationextension = rep(NA, length(citations.funcat.go$geneid)),
                          geneproductformid = rep(NA, length(citations.funcat.go$geneid))
                          )
write.table(funcat.gaf, "./tidy/broad_nc_citations_gofuncat.gaf", sep = "\t", 
            quote = FALSE, row.names = FALSE)

# Need to map GO terms to aspect (B, C or F) (use GO.db package)
library(GO.db)
terms <- as.data.frame(GOTERM)
terms <- terms[, c(2,4,7)]
terms <- transmute(terms, goid1 = go_id, goid2 = Secondary, ontology = Ontology)
terms$ontology <- gsub('.(.)','\\1',terms$ontology )
# join primary and secondary go ids for mapping
terms <- gather(terms, key = go, value = goid, na.rm = TRUE, goid1, goid2)
terms <- terms[,-2]
citations.funcat.go <- read.table("./tidy/broad_nc_citations_gofuncat.gaf",
                                    sep = "\t", header = TRUE, na.strings = c("NA",""))

# Remove leading space in the GO id column
citations.funcat.go$goid <- as.character(citations.funcat.go$goid)
citations.funcat.go$goid <- gsub('\\s','',citations.funcat.go$goid, perl = TRUE)

# Map aspect
citations.funcat.go.gaf <- merge(citations.funcat.go, terms, by.x = "goid", by.y = "goid", all.x = TRUE)
citations.funcat.go.gaf <- distinct(citations.funcat.go.gaf)
citations.funcat.go.gaf$aspect <- citations.funcat.go.gaf$ontology
citations.funcat.go.gaf <- citations.funcat.go.gaf[,c(2,3,4,5,1,6:17)]

# Write to file
write.table(citations.funcat.go.gaf, "./tidy/broad_nc_citations_gofuncat.gaf", sep = "\t", 
            quote = FALSE, row.names = FALSE, na = "")
```

### 3. Gene Ontology Annotations
All GO annotations have the "pubmedid" field empty, but there is a publication reference (authors and title) in the "publication" field. We need the pubmedid for the GAF files

Retrieve pubmedids from Entrez
```{r, eval=FALSE}
# Use rentrez package to recover missing pubmedIDs
library(rentrez)
    
# Filter annotations with no pubmedid
add_pubmedid <- citations.wide[((citations.wide$publication != "None") & (citations.wide$pubmedid == "None")), c("geneid", "publication", "pubmedid")]
    
# Remove html formating tags and keep only the title for searching
add_pubmedid$publication <- gsub("^.*?<i>([^<]+?)</i>","\\1",add_pubmedid$publication, perl = TRUE)

# Inicialize vector to store publications with retrieval errors
nopub <- vector(mode = "integer")

for (i in 1:length(add_pubmedid$publication)) { # test first with 1:100
    if (grepl("^[^\\d]", add_pubmedid$publication[i], perl = TRUE)){
        term <- add_pubmedid$publication[i]
            
        # Remove any NOTs in the title (will be translated to logical NOT by pubmed)
        # Remove other characters/words that are making problems too
        term <- gsub("not", "", term, ignore.case = TRUE, perl = TRUE)
        term <- gsub(" \\w ", " ", term, ignore.case = TRUE, perl = TRUE)
        term <- gsub(" de novo ", " ", term, ignore.case = TRUE, perl = TRUE)
        result_paper <- entrez_search(db="pubmed", term=term )

        if (length(result_paper$ids) != 1){
            # If no result returned, or returned more than one paper, 
            # store index for manual checking later
            nopub <- c(nopub, i)
    
        }else{ # add pubmedid to table
            add_pubmedid$pubMedId[i] <- result_paper$ids
        }
    
    }else if(grepl("^(\\d){1,8}$", add_pubmedid$publication[i], perl = TRUE) ){
        # If there is a pubmedid in the publication field, transfer it to the pubmedid field
        add_pubmedid$pubMedId[i] <- add_pubmedid$publication[i]
    }
}
# temporally store the downloaded pubmeids 
write.table(add_pubmedid, "./data/add_pubmedid_temp.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
```

Check the publications with retrieval errors
```{r, eval = FALSE}
addpubmedid.table <- data.table(add_pubmedid)
missingpubs.table <- data.table(unique(addpubmedid.table[nopub,], by = "publication")) # (106 missing)
write.table(select(missingpubs.table, publication, pubmedid), "./data/missingpubmedids_temp.tsv", sep = "\t", row.names = FALSE, quote = FALSE)
# Load manually added pubmedids (93/106)
missingpubsmanual <- read.table("./data/missingpubmeids_manual.tsv", sep = "\t", header = TRUE, colClasses = c("character","character"), fill = TRUE, quote = "")

# Map manual pubmedids to the list of publications 
for (i in 1:length(missingpubsmanual$publication)){
    indx <- which(missingpubsmanual$publication[[i]] == addpubmedid.table$publication)
    addpubmedid.table[indx, "pubmedid"] <- missingpubsmanual$pubmedid[[i]]
}
```

Add the missing pubmedids to publications in the citations.wide table
```{r, eval = FALSE}
# Simplify addpubmedid.table, reduce it to unique entries that have pubmedids (faster mapping later)
addpubmedid.unique.table <- unique(addpubmedid.table, by = "publication")
addpubmedid.unique.table <- addpubmedid.unique.table[addpubmedid.unique.table$pubmedid != "None",]

# Map pubmedids to publications (in citations.wide table) 
for (i in 1:length(addpubmedid.unique.table$publication)){
#for (i in 1:100){ # Weird bug, some pubmedids are not being replaced with the 1:length loop
    indx <- which(grepl(addpubmedid.unique.table$publication[[i]], 
                        citations.wide$publication, fixed = TRUE))
    citations.wide[indx, "pubmedid"] <- addpubmedid.unique.table$pubmedid[[i]]
    print(i)
}
```

Parse GO term annotations to GAF format
```{r, eval = FALSE}
# Filter GO term annotations (6479)
citations.goterms <- citations.wide[citations.wide$termontology == "Gene Ontology",                                    c("geneid","evidencecode","pubmedid","publication","termid","termname","termnamespace")] 

# Evidence code, only the abbreviation needed
citations.goterms$evidencecode <- gsub("^(.+?)\\/.+$", "\\1", 
                                        citations.goterms$evidencecode, perl = TRUE)
# Remove leading space in the GO id column
citations.goterms$termid <- gsub('\\s','',citations.goterms$termid, perl = TRUE)
# Aspect, abbreviation needed (C, F or B)
citations.goterms$termnamespace <- gsub("cellular_component", "C", 
                                         citations.goterms$termnamespace, perl = TRUE)
citations.goterms$termnamespace <- gsub("biological_process", "P", 
                                         citations.goterms$termnamespace, perl = TRUE)
citations.goterms$termnamespace <- gsub("molecular_function", "F", 
                                         citations.goterms$termnamespace, perl = TRUE)
# Remove html formating tags
citations.goterms$publication <- gsub("<br/><i>([^<]+?)</i>","\\. \\1",citations.goterms$publication, 
                                      perl = TRUE)
# Check GO annotations with no pubmeids, besides the ISS ones
nopublications <- citations.goterms[citations.goterms$pubmedid == "None" &
                                        citations.goterms$evidencecode != "ISS", ]
# Remove annotations with ISS as evidence code and no withfrom or pubmedid
citations.goterms <- filter(citations.goterms, !(citations.goterms$pubmedid == 'None' &
                                citations.goterms$evidencecode == 'ISS'))

# Also, check the assignedby and dbobjectype fields
goterms.gaf <- data.frame(db = rep("EuPathDB", length(citations.goterms$geneid)), 
                          dbobjectid = citations.goterms$geneid, # geneid
                          dbobjectsymbol = rep(NA, length(citations.goterms$geneid)), # gene name
                          qualifier = rep(NA, length(citations.goterms$geneid)),
                          goid = citations.goterms$termid,
                          dbreference = paste("PMID:", citations.goterms$pubmedid, sep = ""),
                          evidencecode = citations.goterms$evidencecode,
                          withfrom = rep(NA, length(citations.goterms$geneid)), # none
                          aspect = citations.goterms$termnamespace,
                          dbobjectname = rep(NA, length(citations.goterms$geneid)), # product
                          dbobjectsynonym = rep(NA, length(citations.goterms$geneid)), # name aliases
                          dbobjecttype = rep("gene", length(citations.goterms$geneid)),
                          taxon = rep("taxon:367110", length(citations.goterms$geneid)),
                          date = rep("20140201", length(citations.goterms$geneid)),
                          assignedby = rep("Broad_NEUROSPORA", length(citations.goterms$geneid)),
                          annotationextension = rep(NA, length(citations.goterms$geneid)),
                          geneproductformid = rep(NA, length(citations.goterms$geneid))
                          )
# Add required GO_REF to annotations with no publication associated
goterms.gaf$dbreference <- gsub("PMID:None", "GO_REF:0000001", goterms.gaf$dbreference)
# Write to GAF
write.table(goterms.gaf, "./tidy/broad_nc_citations_goterms.gaf", sep = "\t", quote = FALSE, 
            row.names = FALSE, na = "")
```

Write also to tab file for reference
```{r, eval = FALSE}
write.table(citations.goterms, "./tidy/citations_goterms.tsv", sep = "\t", quote = FALSE, 
            row.names = FALSE)
```

### 4. Publications
Filter termontology entries type "None" (Publications)
```{r, eval=FALSE}
citations.pub <- citations.wide[citations.wide$termontology == "None", 
                                c("geneid", "pubmedid", "publication")] # (1134)
citations.pub$annotationmethod <- rep("Manual", length(citations.pub$geneid))
citations.pub$annotationsource <- rep("Broad_CAP_NC12_7", length(citations.pub$geneid))


# Remove html formating tags
citations.pub$publication <- gsub("<br/><i>([^<]+?)</i>","\\. \\1",citations.pub$publication, 
                                      perl = TRUE)

# Check for references that are missing the PubMedIDs (OK)
indx <- grep("^[^(\\d)]", citations.pub$pubmedid, perl = TRUE)
citations.pub.noids <- citations.pub[indx, ]
length(citations.pub.noids$publication) # (20/4134 OK)

# Remove entries with no pubmedid and write publication file for loading
indx <- grep("^\\d", citations.pub$pubmedid, perl = TRUE)
citations.pub <- citations.pub[indx,]
write.table(citations.pub, file = "./tidy/citations_publications.tsv", quote = FALSE, 
            sep = "\t", row.names = FALSE)

# Write publications with no PubmedID for reference
write.table(citations.pub.noids, file = "./tidy/citations_publications_nopubmedid.tsv", 
            quote = FALSE, sep = "\t", row.names = FALSE)
```

## ToDo
- [X] Add aspect to Functional Catalogue GO term annotations 