---
title: "Neurospora Broad GeneID Mappings"
output: html_document
---

## Intro

Generate tab files to load previous ids into FungiDB, one entry per row. There is an excel file with mappings between the different versions at:

[Broad dump](http://fungalgenomes.org/public/neurospora/data/Broad_dump/Dump_2014_02.tar.gz)

It also contains comments about the type of changes.

Dumped all the tabs from the excel file as csv files:

[dumped tab files]("./data/Neurospora_crassa_versions_geneid_mappings/")

## Create individual mapping files for loading and easy reference

### v1 mapping

Only one pair old:new per row in the source file (merges and splits are unnested in the file)

```{r, eval = FALSE}
v1mappings <- read.table("./data/Neurospora_crassa_versions_geneid_mappings/v1-to-v7-Table 1.csv",
                         quote = "", header = TRUE, strip.white = TRUE, sep = ",")

# Filter only v1 and v7 IDs, and remove version from the current geneIDs (v7)
v1mappings <- transmute(v1mappings, geneid = gsub("^(.+)\\.7","\\1", v1mappings$NC12_CALLGENES_FINAL_7,
                            perl = TRUE), geneidv1 = I(v1mappings$NC3_FINAL_CALLGENES_1))

# Remove added and deleted genes (no mappings)
v1mappings <- filter(v1mappings, geneid != "")
v1mappings <- filter(v1mappings, geneidv1 != "")

# Dump mappings v7 to v1
write.table(v1mappings, "./tidy/broad_nc_geneid_mappings_v1.tsv", row.names = FALSE, 
            quote = FALSE, sep = "\t")
```

### v2 mapping

Only one pair old:new per row in the source file (merges and splits are unnested in the file)

```{r, eval = FALSE}
v2mappings <- read.table("./data/Neurospora_crassa_versions_geneid_mappings/v2-to-v7-Table 1.csv",
                         quote = "", header = TRUE, strip.white = TRUE, sep = ",", comment.char = "",
                         row.names = NULL)

# Filter only v2 and v7 IDs, and remove version from the current geneIDs (v7)
v2mappings <- transmute(v2mappings, geneid = gsub("^(.+)\\.7","\\1", v2mappings$NEW_LOCUS,
                                                  perl = TRUE), 
                        geneidv2 = I(v2mappings$OLD_LOCUS))

# Remove added and deleted genes (no mappings)
v2mappings <- filter(v2mappings, geneid != "DELETE")
v2mappings <- filter(v2mappings, geneid != "ADD") # geneids where in the wrong column anyway

# Remove also "notmapped"... I guess they were deleted too (they are not in v7)
v2mappings <- filter(v2mappings, geneid != "NOTMAPPED")

write.table(v2mappings, "./tidy/broad_nc_geneid_mappings_v2.tsv", row.names = FALSE, 
            quote = FALSE, sep = "\t")
```

### v3 mappings

Only one pair old:new per row in the source file (merges and splits are unnested in the file)

```{r, eval = FALSE}
v3mappings <- read.table("./data/Neurospora_crassa_versions_geneid_mappings/v3-to-v7-Table 1.csv",
                         quote = "", header = TRUE, strip.white = TRUE, sep = ",", comment.char = "",
                         row.names = NULL)

# Filter only v3 and v7 IDs, and remove version from the current geneIDs (v7)
v3mappings <- transmute(v3mappings, geneid = gsub("^(.+)\\.7","\\1", v3mappings$NC12_CALLGENES_FINAL_7,
                                                  perl = TRUE), 
                        geneidv3 = I(v3mappings$NC7_FINAL_CALLGENES_3))

# Remove added and deleted genes (no mappings)
v3mappings <- filter(v3mappings, !(geneid == "" | geneidv3 == ""))

write.table(v3mappings, "./tidy/broad_nc_geneid_mappings_v3.tsv", row.names = FALSE, 
            quote = FALSE, sep = "\t")
```

### v4 mappings

There are 2 files for v4 to v7 mappings in the Broad dump, using the latest (Jun vs Feb)

Merges and split are nested in the source file, they will need unnesting.

There is a split with the new gene loci "unconfirmed", but there is also a note in the current annotation confirming the split gene IDs

```{r, eval = FALSE}
v4mappings <- read.table("./data/Neurospora_crassa_versions_geneid_mappings/v4_jun-to-v7-Table 1.csv",
                         quote = "", header = TRUE, strip.white = TRUE, sep = ",", comment.char = "",
                         row.names = NULL, skip = 1)

# Filter only v4 and v7 IDs, no version attached to the geneIDs
v4mappings <- transmute(v4mappings, geneid = as.character(v4mappings$NC12_CALLGENES_FINAL_7_locus),
                        geneidv4 = as.character(v4mappings$NC10_CALLGENES_FINAL_4_JUN_locus) )

# Add missing split IDs
v4mappings$geneid[which(v4mappings$geneidv4 == "NCU07571")] <- "NCU07571/NCU15832"

# Remove added and deleted genes (no mappings)
v4mappings <- filter(v4mappings, !(geneid == "" | geneidv4 == ""))

# Unnest merges and splits

v4mappings <- transmute(v4mappings, geneid = strsplit(as.character(geneid), '/'),
                        geneidv4 = strsplit(as.character(geneidv4), '/') )

v4mappings <- unnest(v4mappings, geneid, .drop = FALSE )
v4mappings <- unnest(v4mappings, geneidv4, .drop = FALSE )

# Add version tag to v4 gene IDs
v4mappings$geneidv4 <- paste0(v4mappings$geneidv4,".4")

# Write to file
write.table(v4mappings, "./tidy/broad_nc_geneid_mappings_v4.tsv", row.names = FALSE, 
            quote = FALSE, sep = "\t")
```

### v5 mappings

v5 file looks like the file for v4 mappings, but it is not the same (checked). 

Some gene IDs where reused between v4 and v5, careful with the mapping... you need to know the version to map some IDs (ex: NCU11858.4 is NCU00033, but NCU11858.5 is NCU16695)

Merges and split are nested in the source file, they will need unnesting.

There is a split with the new gene loci "unconfirmed", but there is also a note in the current annotation confirming the split gene IDs

```{r, eval = FALSE}
v5mappings <- read.table("./data/Neurospora_crassa_versions_geneid_mappings/v5-to-v7-Table 1.csv",
                         quote = "", header = TRUE, strip.white = TRUE, sep = ",", comment.char = "",
                         row.names = NULL, skip = 1)

# Filter only v5 and v7 IDs, no version attached to the geneIDs
v5mappings <- transmute(v5mappings, geneid = as.character(v5mappings$NC12_CALLGENES_FINAL_7_LOCUS_ID),
                        geneidv5 = as.character(v5mappings$NC10_CALLGENES_FINAL_5_locus) )

# Add missing split IDs
v5mappings$geneid[which(v5mappings$geneidv5 == "NCU07571")] <- "NCU07571/NCU15832"

# Remove added and deleted genes (no mappings)
v5mappings <- filter(v5mappings, !(geneid == "" | geneidv5 == ""))

# Unnest merges and splits

v5mappings <- transmute(v5mappings, geneid = strsplit(as.character(geneid), '/'),
                        geneidv5 = strsplit(as.character(geneidv5), '/') )

v5mappings <- unnest(v5mappings, geneid, .drop = FALSE )
v5mappings <- unnest(v5mappings, geneidv5, .drop = FALSE )

# Add version tag to v5 gene IDs

v5mappings$geneidv5 <- paste0(v5mappings$geneidv5,".5")

# Write to file
write.table(v5mappings, "./tidy/broad_nc_geneid_mappings_v5.tsv", row.names = FALSE, 
            quote = FALSE, sep = "\t")

# Compare v4 to v5 IDs (just checking if they are the same, they are not)
mappingstest <- merge(v4mappings, v5mappings, by.x = "geneid", by.y = "geneid")
```


Create a combined file, with all the mappings, one column per version
```{r, eval = FALSE}
mappingall <- merge(v1mappings, v2mappings,  by.x = "geneid", by.y = "geneid", all = TRUE)
mappingall <- merge(mappingall, v3mappings,  by.x = "geneid", by.y = "geneid", all = TRUE)
mappingall <- merge(mappingall, v4mappings,  by.x = "geneid", by.y = "geneid", all = TRUE)
mappingall <- merge(mappingall, v5mappings,  by.x = "geneid", by.y = "geneid", all = TRUE)

# Write to file
write.table(mappingall, "./tidy/broad_nc_geneid_mappings_all.tsv", row.names = FALSE, 
            quote = FALSE, sep = "\t")
```

Create also a combined file, with all the mappings in the same column
```{r, eval = FALSE}
mappingone <- gather(mappingall, na.rm = TRUE, key = version, value = geneidv, geneidv1:geneidv5)
# Get read of the version column
mappingone <- mappingone[,-2]
# Sort by gene ID
mappingone <- mappingone[order(mappingone$geneid),]

# Write to file
write.table(mappingone, "./tidy/broad_nc_geneid_mappings_all_one_column.tsv", row.names = FALSE,
            quote = FALSE, sep = "\t")
```